---
title: "Multitrait BLUP with metan"
author: "Maarit Mäenpää"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

A case of a single experiment: One environment, multiple genotypes.

### Packages needed in the example

```{r include=T, echo=T, message=F, warning=F}

library(agridat) # for dataset
library(dplyr)
library(metan)
library(ggplot2)

```

# Prepare an example dataset

Using barrero.maize from agridat. This is a multi-year, multi-trait, multi-environment dataset, where for the sake of an example, we extract one year and one location. We omit all missing values.

```{r include=T, echo=T}

# Load in the data
data(barrero.maize)
head(barrero.maize)

# For details of the data
#?barrero.maize

# Select one location in one year
maize.BA <- filter(barrero.maize, loc=="BA" & year=="2000")

# Omit all NA's
maize.BA <- na.omit(maize.BA)

# Make sure genotypes that are removed due to missing data are not stored as a level in the genotype by reassigning the factor
maize.BA$gen <- factor(maize.BA$gen)

```

# Run a model with metan

Need to specify the dataset (maize.BA), the column identifying the genotype (gen), replication (rep), and the response variables of interest. Here, there are 8 traits measured. 

```{r include=T, echo=T, warning=FALSE, message=FALSE}

mod <-
  gamem(maize.BA, 
        gen=gen, 
        rep=rep, 
        resp=c(daystoflower, plantheight, earheight, population, lodged, moisture, testweight, yield))

```

Note that the output after running the model indicates a brief overview of which traits differ between genotypes. In this case only the trait "lodged" did not differentiate between different genotypes.


# BLUPs


Plot the individual BLUPs (best linear unbiased predictors) from the model. 

```{r include=T, echo=T}

plot_blup(mod, var = "daystoflower")
plot_blup(mod, var = "plantheight")
plot_blup(mod, var = "earheight")
plot_blup(mod, var = "population")
plot_blup(mod, var = "lodged")
plot_blup(mod, var = "moisture")
plot_blup(mod, var = "testweight")
plot_blup(mod, var = "yield")

```

# Assess the model

### Broad sense heritability

Proportion of phenotypic variance due to genetic factors.

```{r include=T, echo=T}
get_model_data(mod, "h2") # broad-sense heritability
```

### Model details

**Ngen** - The number of genetic entries or genotypes evaluated.

**OVmean** - The overall mean value for each trait across all genotypes.

**Min** - The minimum observed value for each trait, with the genotype and replication noted.

**Max** - The maximum observed value for each trait, with the genotype and replication noted.

**MinGEN** - The minimum genetic value for each trait, indicating the lowest value among the genotypes.

**MaxGEN** - The maximum genetic value for each trait, indicating the highest value among the genotypes.

```{r include=T, echo=T}

get_model_data(mod, "details") # General details
```
### Likelihood ratio tests

These tests give the p-values (remember, p-values are not everything!) for whether the traits assessed show differences in the genotypes. 

```{r include=T, echo=T}
get_model_data(mod, "lrt") # P-values for genotype differences within traits
```

### Genetic parameters 

**Gen_var** - Genetic variance, indicating the variability in traits due to genetic differences.

**Gen (%)** - Percentage of total variance attributed to genetic factors.

**Res_var** - Residual variance, representing the variability not explained by genetic or block effects.

**Res (%)** - Percentage of total variance attributed to residual factors.

**Phen_var** - Phenotypic variance, the total observed variability in the traits.

**H2** - Broad-sense heritability, indicating the proportion of phenotypic variance due to genetic factors.

**h2mg** - Narrow-sense heritability, showing the proportion of phenotypic variance due to additive genetic factors.

**Accuracy** - The accuracy of the heritability estimates.

**CVg** - Coefficient of genetic variation, a measure of genetic diversity relative to the mean.

**CVr** - Coefficient of residual variation, indicating the residual variability relative to the mean.

**CV ratio** - Ratio of genetic to residual variation, providing insight into the relative contributions of genetic and residual factors

```{r include=T, echo=T}
get_model_data(mod, "genpar") # Genetic parameters
```

### BLUPs

I.e. predicted means (=point estimates for genotype mean in each trait)

```{r include=T, echo=T}
get_model_data(mod, "blupg") # Predicted means

```

# FAI-index

Multitrait comparison. The method is documented first here: https://doi.org/10.1111/gcbb.12443

```{r include=T, echo=T}

FAI <- fai_blup(mod)

plot(FAI)

```

## FAI ranking and illustrating selected genotypes

Here, genotype selection is not purely for the optimal top performing genotypes, but for diversifying the best- and worst performing genotypes. In this case, we take the top and bottom 2 genotypes, and randomly choose 4 from the middle. 


```{r include=T, echo=T}

# Save BLUP data for individual traits
metan_blups <- data.frame(get_model_data(mod, "ranef")$GEN)

# Extract FAI indexes
FAI_results <- FAI$FAI

# Change the name of Genotype to combine the two dataframes
names(metan_blups)[1] <- "Genotype"

# Combine dataframes
FAI_results <- left_join(FAI_results, metan_blups, by="Genotype")

# Create a dataframe to select genotypes
selected <- data.frame(matrix(NA, nrow = 0, ncol = 25))
colnames(selected) <- colnames(FAI_results)

# Top and bottom 5
selected <- rbind(selected, 
                  FAI_results[which(FAI_results$ID1 %in% tail(sort(FAI_results$ID1),2)==T),]
)
selected <- rbind(selected, 
                  FAI_results[which(FAI_results$ID1 %in% head(sort(FAI_results$ID1),2)==T),]
)

# Random selection from the middle
rselect <- sample(FAI_results[which(FAI_results$Genotype %in% selected$Genotype==FALSE),]$Genotype, size = 4, replace = F)
selected <- rbind(selected, FAI_results[which(FAI_results$Genotype %in% rselect ==T),])

# Illustrate selection
ggplot(FAI_results, aes(x=ID2, y=ID1)) + 
  geom_point() + 
  geom_point(data=selected, colour="red")


```

The figure above illustrates the ranked order of the genotypes. The red are selected, and black are not. Distance in ID1 and ID2 illustrate difference between the genotypes. 
